---
title: "Lab 05 - Wrangling spatial data"
author: "Cynthia Jiao"
date: "Insert date here"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox) 
```

```{r load-data, message = FALSE}
states <- read_csv("data/states.csv")
```

### Exercise 1 & 2

```{r}
## filtering out all Dennys and La quinta in AK

dn_ak <- dennys %>%
  filter(state == "AK")
nrow(dn_ak)

lq_ak <- laquinta %>%
  filter(state == "AK")
nrow(lq_ak)

print(nrow(dn_ak)*nrow(lq_ak))

```
There are 3 Denny's and 2 La quinta in Alaska. And there are 6 combinations of distance that we need to calculate. 

### Exercise 3 & 4

```{r}
## combing two AK dataset

dn_lq_ak <- full_join(dn_ak, lq_ak, 
                      by = "state")
print(dn_lq_ak)

```
There are 6 observations in the new data frame. There are 13 variables as the two original dataset were joined by "state." All rest variables are doubled and marked as .x or .y.

### Exercise 5

```{r}

## creating the Haversine function

haversine <- function(long1, lat1, long2, lat2, round = 3) {
  # convert to radians
  long1 <- long1 * pi / 180
  lat1 <- lat1 * pi / 180
  long2 <- long2 * pi / 180
  lat2 <- lat2 * pi / 180

  R <- 6371 # Earth mean radius in km

  a <- sin((lat2 - lat1) / 2)^2 + cos(lat1) * cos(lat2) * sin((long2 - long1) / 2)^2
  d <- R * 2 * asin(sqrt(a))

  return(round(d, round)) # distance in km
}


```

### Exercise 6

```{r}

## using Haversine function to calculate distances

dn_lq_ak <- dn_lq_ak %>%
  mutate(distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y, round = 3))


```

### Exercise 7

```{r}
## calculate the minimum distance between a Denny’s and La Quinta for each Denny’s location

dn_lq_ak_mindist <- dn_lq_ak %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))

print(dn_lq_ak_mindist)

```

### Exercise 8

There are 3 Denny's locations in Alaska that are nearest to a La quinta location. Out of the 6 locations, 3 Denny’s locations in Alaska have a nearby La Quinta within a relatively short distance (under 7 km), indicating that these two chains tend to be co-located in urban areas. However, with only three data points, it's difficult to generalize broader patterns.
Out of 3 nearest locations, min distance = 2.04, max distance = 6, mean distance = 4.41, median distance = 5.2, sd = 2.09
```{r}
## bar plot for the distance between Denny's to Nearest La Quinta for the 3 locations in Alaska.

ggplot(dn_lq_ak_mindist, aes(x = address.x, y = closest)) +
  geom_bar(stat = "identity", fill = "purple", alpha = 0.5) +
  labs(title = "Distance from Denny's to Nearest La Quinta (Alaska)",
       x = "Denny's locations",
       y = "Distance") +
  theme_minimal()

## summary stats 

summary_stats_ak <- dn_lq_ak_mindist %>%
  summarize(
    mean_distance = mean(closest),
    median_distance = median(closest),
    sd_distance = sd(closest),
    min_distance = min(closest),
    max_distance = max(closest),
    iqr_distance = IQR(closest)
  )

print(summary_stats_ak)

```

### Exercise 9

Repeat the same analysis for North Carolina
```{r}

## filtering out all Dennys and La quinta in NC

dn_nc <- dennys %>%
  filter(state == "NC")
nrow(dn_nc)

lq_nc <- laquinta %>%
  filter(state == "NC")
nrow(lq_nc)

print(nrow(dn_nc)*nrow(lq_nc))

```

There are 28 Denny's locations in NC that are nearest to a La quinta location. The distribution of distances between Denny’s and the nearest La Quinta locations in North Carolina is highly variable, ranging from near zero to over 150 km. Under the distance of 25km, there are 8 Denny's location that have a La quinta nearby.
Between the closest Denny's & La quinta locations in NC, min distance = 1.78, max distance = 188, mean distance = 65.4, and median distance = 53.5, sd = 53.4
```{r}

## combing two NC dataset

dn_lq_nc <- full_join(dn_nc, lq_nc, 
                      by = "state")
print(dn_lq_nc)

## using Haversine function to calculate distances

dn_lq_nc <- dn_lq_nc %>%
  mutate(distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y, round = 3))

## calculate the minimum distance between a Denny’s and La Quinta for each Denny’s location

dn_lq_nc_mindist <- dn_lq_nc %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))

print(dn_lq_nc_mindist)

## bar plot for the distance between Denny's to Nearest La Quinta for the 3 locations in NC

ggplot(dn_lq_nc_mindist, aes(x = address.x, y = closest)) +
  geom_bar(bindwidth = 10, stat = "identity", fill = "purple", alpha = 0.5) +
  labs(title = "Distance from Denny's to Nearest La Quinta (North Carolina)",
       x = "Denny's locations",
       y = "Distance") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) 

## summary stats 

summary_stats_nc <- dn_lq_nc_mindist %>%
  summarize(
    mean_distance = mean(closest),
    median_distance = median(closest),
    sd_distance = sd(closest),
    min_distance = min(closest),
    max_distance = max(closest),
    iqr_distance = IQR(closest)
  )

print(summary_stats_nc)

```

### Exercise 10

Repeat the same analysis for Texas
```{r}

## filtering out all Dennys and La quinta in TX

dn_tx <- dennys %>%
  filter(state == "TX")
nrow(dn_tx)

lq_tx <- laquinta %>%
  filter(state == "TX")
nrow(lq_tx)

print(nrow(dn_tx)*nrow(lq_tx))

```

There are 200 Denny's locations in TX that are nearest to a La quinta location. Most distances concentrated below 20 km, but several locations extend beyond 40 km. The histogram suggests that many Denny’s are relatively close to La Quinta hotels locations, but there are also some outliers, where the nearest La Quinta is up to 60 km far away.
Between the closest Denny's & La quinta locations in TX, min distance = 0.016, max distance = 60.1, mean distance = 5.8, and median distance = 3.37, sd = 8.8
```{r}

## combing two TX dataset

dn_lq_tx <- full_join(dn_tx, lq_tx, 
                      by = "state")
print(dn_lq_tx)

## using Haversine function to calculate distances

dn_lq_tx <- dn_lq_tx %>%
  mutate(distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y, round = 3))

## calculate the minimum distance between a Denny’s and La Quinta for each Denny’s location

dn_lq_tx_mindist <- dn_lq_tx %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))

print(dn_lq_tx_mindist)

## bar plot for the distance between Denny's to Nearest La Quinta for the 3 locations in NC

ggplot(dn_lq_tx_mindist, aes(x = address.x, y = closest)) +
  geom_bar(bindwidth = 10, stat = "identity", fill = "purple", alpha = 0.5) +
  labs(title = "Distance from Denny's to Nearest La Quinta (Texas)",
       x = "Denny's locations",
       y = "Distance") +
  theme_minimal() + 
  theme(axis.text.x = element_blank())  ## remove the exact address of each location...


## summary stats 

summary_stats_tx <- dn_lq_tx_mindist %>%
  summarize(
    mean_distance = mean(closest),
    median_distance = median(closest),
    sd_distance = sd(closest),
    min_distance = min(closest),
    max_distance = max(closest),
    iqr_distance = IQR(closest)
  )

print(summary_stats_tx)


```

### Exercise 11

Repeat the same analysis for New York
```{r}

## filtering out all Dennys and La quinta in NY

dn_ny <- dennys %>%
  filter(state == "NY")
nrow(dn_ny)

lq_ny <- laquinta %>%
  filter(state == "NY")
nrow(lq_ny)

print(nrow(dn_ny)*nrow(lq_ny))

```

There are 200 Denny's locations in NY that are nearest to a La quinta location. The distribution of distances between Denny’s and the nearest La Quinta locations in New York is highly variable, ranging from near zero up to 100km.
Between the closest Denny's & La quinta locations in NY, min distance = 1.28, max distance = 99, mean distance = 33.6, and median distance = 24.158, sd = 28.7
```{r}

## combing two NY dataset

dn_lq_ny <- full_join(dn_ny, lq_ny, 
                      by = "state")
print(dn_lq_ny)

## using Haversine function to calculate distances

dn_lq_ny <- dn_lq_ny %>%
  mutate(distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y, round = 3))

## calculate the minimum distance between a Denny’s and La Quinta for each Denny’s location

dn_lq_ny_mindist <- dn_lq_ny %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))

print(dn_lq_ny_mindist)

## bar plot for the distance between Denny's to Nearest La Quinta for the 3 locations in NY

ggplot(dn_lq_ny_mindist, aes(x = address.x, y = closest)) +
  geom_bar(bindwidth = 10, stat = "identity", fill = "purple", alpha = 0.5) +
  labs(title = "Distance from Denny's to Nearest La Quinta (New York)",
       x = "Denny's locations",
       y = "Distance") +
  theme_minimal() +
  theme(axis.text.x = element_blank())  ## remove the exact address of each location...


## summary stats 

summary_stats_ny <- dn_lq_ny_mindist %>%
  summarize(
    mean_distance = mean(closest),
    median_distance = median(closest),
    sd_distance = sd(closest),
    min_distance = min(closest),
    max_distance = max(closest),
    iqr_distance = IQR(closest)
  )

print(summary_stats_ny)


```


### Exercise 12

Among the four states examined, Texas is where Mitch Hedberg’s joke is most likely to hold true, because of its relatively small standard deviation of distance between Denny's and La quinta. The majority of Denny’s locations in Texas have a nearby La Quinta within 20 km, with relatively few exceeding 40 km. In contrast, North Carolina and New York show much greater variability in their closest combinations, with many locations over 50–100 km apart, making the joke less applicable. In addition, although Alaska has the shortest absolute distances, its small sample size limits generalization. 


